\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-27 analysis/manuscript.tex
final glitches

By the end of 2023, many transformative agreements had been implemented,
but the interim outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both fully open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilisation of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements has contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuates market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Klebel \& Ross-Hellauer, 2023), because transformative
agreements' focus on pay to publish mainly targets institutions from
high-income countries, furthering a questionable journal prestige
culture (Babini et al., 2022). Besides, an editorial-board resignation
raised concerns that transformative agreements might encourage
publishers to maximize journal publication volume ``without regard to
quality'' (Rasmussen, 2023).

@@

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). In particular, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements. Likewise, article-level open access invoicing
data are only partly available (Jahn et al., 2022).

@@

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, the steps are described
in detail.

@@

Comparing Crossref license coverage with OpenAlex, which re-uses open
access evidence from Unpaywall, a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), highlighted ongoing challenges in identifying
hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
under CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The largest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers differ only slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable differences
could be observed among some publishers that presumably did not provide
CC license information to Crossref, including AIP Publishing, the
American Physiological Society, Emerald, and the Royal Society. Crossref
license metadata was more complete with regard to the articles published
by Wiley and the American Chemical Society. Finally, inconsistent open
access status information in previous OpenAlex versions was observed
(Jahn et al., 2023). After reporting this to OpenAlex, fixing this issue
was still ongoing according to the release notes, which might also
explain part of the discrepancy.

@@

Throughout this mostly automated data gathering and analysis process,
Tidyverse tools (Wickham et al., 2019) for the R programming language (R
Core Team, 2020) were used. The resulting data are openly available
through an R data package, hoaddata, version 0.2.91 Following Marwick et
al. (2018), hoaddata contains not only the datasets used in the data
analysis. It also includes code used to compile the data by connecting
it to the cloud-based Google Big Query data warehouse, where scholarly
big data from Crossref, OpenAlex and Unpaywall were imported. To
increase computational reproducibility, data aggregation through
hoaddata was automatically carried out using GitHub Actions, a
continuous integration service.

@@

Between 2018 and 2022, 11,189 out of 12,857 hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 of the 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). Simultaneously,
the total article volume of the investigated journals increased from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable increase in open
access provision through other means, presumably publication fees being
not invoiced through transformative agreements, which increased from
4.1\% (n = 62,625) in 2018 to 6.3\% (n = 105,896).

@@

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, started offering open access
agreements for its hybrid journal portfolio to selected consortia such
as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish
FinELib consortium in 2015 under the name Springer Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
Prior to this, Springer discontinued its hybrid OA pilot agreements with
a small number of institutions, including the University of Göttingen
(Schmidt \& Shearer, 2012). However, the Springer Compact agreements
were not included in the data as they concluded prior to the start of
the transformative agreement data collection in June 2021. Nonetheless,
the results demonstrate the importance of agreements for Springer
Nature's hybrid open access business over the past five years (Figure
2B). In 2022, 66\% (n = 34,725) of open access in Springer Nature hybrid
journals was enabled through transformative agreements. In the same
year, 70\% (n = 37,316) of Wiley's open access articles could be linked
to transformative agreements in 2022. By contrast, Elsevier published
fewer than half of its open access articles through transformative
agreements (n = 32,627; 54\%).

@@

\caption{\label{tab:subject_summary_table}Hybrid open access through transformative agreements by journal subject 2018-2022}
\centering
\begin{tabular}[t]{lrlrlrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hybrid journals} & \multicolumn{2}{c}{Articles} & \multicolumn{2}{c}{OA articles} & \multicolumn{2}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Journal subject & Total & \% & Total & \% & Total & \% & Total & \%\\
\midrule
Health Sciences & 2,376 & 22.5 & 2,709,906 & 27.8 & 286,592 & 27.3 & 117,746 & 25\\
Life Sciences & 1,403 & 13.3 & 1,477,808 & 15.1 & 191,880 & 18.3 & 71,593 & 15.2\\
Physical Sciences & 2,732 & 25.9 & 4,291,833 & 44 & 366,794 & 35 & 167,686 & 35.6\\
Social Sciences & 4,050 & 38.3 & 1,280,460 & 13.1 & 203,461 & 19.4 & 114,190 & 24.2\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018-2022. In particular, Social
Sciences inbcluding Arts and Humanties journals accounted for the
strongest growth in the five-years period from 6.4\% (n = 8,361) to 23\%
(n = 51,938), followed by the Life Sciences from 7.6\% (n = 15,003) to
18\% (n = 39,494) , Health Sciences from 5.3\% (n = 18,279) to 16\% (n =
63,089) and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n =
85,428). Growth in Social Sciences category can be largely attributed to
transformative agreements. In 2022, two-thirds of open access articles
(67\%, n = 34,759) were published by the first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B).

@@

Figure \ref{fig:subject_panel}C shows that this trend was consistent
across hybrid journals belonging to the ASJC Social Sciences category.
In 2022, 25\% of Social Sciences journals provided open access to at
least every fourth article exclusively through transformative
agreements. However, hybrid open access through transformative
agreements played a comparably lesser role in Life Sciences and Health
Sciences. In these two subject areas, only about half of the open access
articles can be linked to these agreements, both overall and on a median
average across journals. In contrast, the majority of Physical Science
Journals show an increase of open access through transformative
agreements compared to other options to publish open access in hybrid
journals.

@@

Between 2018 and 2022, high-income countries almost exclusively
dominated hybrid open access publishing through transformative
agreements. During this period, first-authors affiliated with
institutions from Organisation for Economic Co-operation and Development
(OECD) member countries published 602,050 open access articles in hybrid
journals, representing 81\% of the investigated open access articles.
This disparity between OECD nations and other countries becomes even
more evident when considering open access through transformative
agreements, as 310,712 of 328,957, or 94\% of open access articles were
associated with such agreements.

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by country, comparing the OECD area with the BRICS, an
intergovernmental organisation, which comprised Brazil, Russia, India,
China, and South Africa as of 2022. The residual category ``Other''
includes the remaining countries. Full counting was applied to account
for multiple institutional affiliations (Hottenrott et al., 2021). From
2018 to 2022, the proportion of open access in hybrid journals increased
from 6.1\% in 2018 to 26\% in 2022. In contrast, BRICS recorded a low
uptake, moderately growing from 1.6\% in 2018 to 3.7\% in 2022.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares the total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (268,965 articles) and India (87,428
articles). In contrast, Western and Northern European countries
published a considerably high number of open access articles.
Particularly, Germany, Great Britain, the Netherlands, Sweden,
Switzerland, and Spain recorded an above-average open access share, as
indicated by the linear trend line. As represented by the point size, as
well as that shown in Figure \ref{fig:country_patch}C, transformative
agreements contributed to these market positions. Interestingly, the
United States had a notable open access market share of 15\%, although
transformative agreements contributed to a lesser extent. Similarly,
China's open access market share of 7.2\% in 2022 was comparable to that
of the Netherlands, which was 7.1\%.

@@

Since 2021, there has been a general trend towards hybrid open access
among many high-income countries, driven primarily by transformative
agreements. However, the proliferation of transformative agreements
differed across these countries. For instance, Germany successfully
negotiated an agreement with Elsevier not until 2023. Additionally,
publication limits or eligibility criteria for institutions and article
types may explain why even countries with widespread agreement
implementation such as Sweden or the Netherlands did not achieve 100\%
hybrid open access. Interestingly, in Japan and the US, options other
than transformative agreements were the main drivers for the increase in
hybrid open access. Once again, the graph highlights countries with low
hybrid open access, particularly non-OECD countries, where only a few or
no agreements were in place.

@@

The primary aim of this study was to assess the uptake of open access in
hybrid journals included in transformative agreements, which were
introduced as a temporal means to support the transition of
subscription-based academic publishing to full open access. This study
presents a novel approach based on open data, which leverages metadata
on over 700 agreements and nine million journal articles to estimate the
extent to which transformation agreements contribute to the transition
of this journal business model. The results highlight a strong growth in
open access between 2018 and 2022, driven by an increasing number of
institutions that had transformative agreements in place. However, the
majority of research literature published in hybrid journals in this
five-year period remained behind the publisher's paywall. Growth in the
adoption of open access in hybrid journals, in particular through
transformative agreements, can be largely attributed to three large
commercial publishers -- Elsevier, Springer Nature, and Wiley -- but
varies substantially across journals, publishers, disciplines, and
country affiliations. Despite the limitations of the data, the findings
indicate that the current level of implementation of transformative
agreements is insufficient to bring about a large-scale transition to
fully open access.

@@

Moreover, this study presents varying levels of open access uptake
through transformative agreements across journals, which can be
attributed to the alignment of authors' affiliations and the
availability of such agreements at their institutions. In line with
previous research findings (Butler et al., 2023; Jahn et al., 2022),
high uptake rates were observed across hybrid journals in Social
Sciences and Humanities. However, it is important to emphasise that
these hybrid journals do not encompass the entire field. For example,
Khanna et al. (2022) found that approximately 60\% of journals utilizing
the open-source publishing platform Open Journal Systems (OJS) fall
within the Social Sciences and Humanities. In these fields, the majority
of full open access journals are so-called ``Diamond OA journals'',
which do not charge publication fees (Bosman, Frantsvåg, et al., 2021).
Rather, this result can be more accurately attributed to the substantial
proportion of authors from high-income countries who publish in these
hybrid journals included in transformative agreements, particularly when
journals are of local relevance, such as by belonging to national
societies or regional research scope.

@@

This large-scale study provides the first empiricial evidence of the
influence of transformative agreements on the transition of hybrid
journals to full open access. However, that study has several
limitations that need to be acknowledged. From a data perspective,
estimations of open access through transformative agreements were
established by linking first author affiliations with publicly available
agreement data from cOAlition S, and not through invoicing data, which
is usually not shared. Moreover, this study is unable to account for the
various types of transformative agreements, particularly with regard to
article types and caps that limit the number of open access articles
covered. Furthermore, assessing the quality of the OpenAlex and Crossref
data used, particularly in terms of affiliations and article types, by
comparing them with established bibliometric databases such as Scopus
and Web of Science was beyond the scope of this analysis. Overall, the
methodology is designed in such a way that the results underestimate
rather than overestimate the adoption of open access through
transformative agreements. In addition, the data used in this study are
openly available, along with the code used for this analysis.

Additionally, it must be noted that the study period was significantly
impacted by the COVID-19 pandemic, which led to an unprecedented number
of publications and a reduction in international collaboration
(Aviv-Reuven \& Rosenfeld, 2021), which could explain the observed
contrasting developments in OECD and BRICS countries. Even before the
pandemic, growth in publications in Europe was only due to
internationally co-authored journal articles (Kwiek, 2021). Likewise,
inflows from China to the US and European countries already declined by
2020 (Zhao et al., 2023). Furthermore, the study design did not consider
emerging publication practices such as preprints (Fraser et al., 2021)
and special issues (Hanson et al., 2023), which have grown rapidly since
2020. Lastly, it should be emphasized that the study did not address
financial shifts between suscriptiuons spending and open acces spayemnts
while analysing hybrid open access through transformative agreements due
to a lack of data on expenditures.

This study allows for multiple strands of further research. One is to
complement this large-scale study with more specific evidence from
individual countries or subjects, particularly those with low hybrid
open access rates. Incorporating full open access and subscription-based
journals, as well as considering global trends in scholarly migration
and collaboration could also be promising. Financial studies could build
upon the study design and incorporate subscription and open access
expenditures to evaluate the cost-effectiveness of transformative
agreements.

@@

In summary, this study provides empirical insights into the development
of hybrid open access following the introduction of transformative
agreements. These results are important for both researchers and
stakeholders engaged in negotiating and evaluating these agreements. The
presented approach relies on open data, which enables follow-up studies
and open access monitoring activties to further explore the role of
transformative agreements in transitioning academic publishing to full
and equitable open access.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Brainard_2023}{}}%
Brainard, J. (2023). {`Transformative'} journals get booted for
switching to open access too slowly. In \emph{Science}.
\url{https://doi.org/10.1126/science.adj3282}

@@

\leavevmode\vadjust pre{\hypertarget{ref-ifla}{}}%
Campbell, C., Dér, Á., Geschuhn, K., \& Valente, A. (2022). How are
transformative agreements transforming libraries? In I. F. of Library
Associations \& I. (IFLA) (Eds.), \emph{87th IFLA world library and
information congress (WLIC) / 2022 in dublin, ireland}. IFLA.
\url{https://repository.ifla.org/handle/123456789/1973}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Geschuhn_2017}{}}%
Geschuhn, K., \& Stone, G. (2017). It's the workflows, stupid! What is
required to make {`offsetting'} work for the open access transition.
\emph{Insights the {UKSG} Journal}, \emph{30}(3), 103--114.
\url{https://doi.org/10.1629/uksg.391}

>>>>

By the end of 2023, many transformative agreements had been implemented,
but the interim outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both fully open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilisation of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements has contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuates market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Ross-Hellauer et al., 2022), because transformative
agreements' focus on pay to publish mainly targets institutions from
high-income countries, furthering a questionable journal prestige
culture (Babini et al., 2022). Besides, an editorial-board resignation
raised concerns that transformative agreements might encourage
publishers to maximize journal publication volume ``without regard to
quality'' (Rasmussen, 2023).

@@

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018; Schmal, 2024), while large-scale
studies relied on self-reported agreement data (Moskovkin et al., 2022),
or used APC pricing lists (Shu \& Larivière, 2023). In particular, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements. Likewise, article-level open access invoicing
data are only partly available (Jahn et al., 2022).

@@

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref (Hendricks et al., 2020) served as the primary
data source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. To determine open access articles published through
transformative agreements, first author affiliations from OpenAlex
(Priem et al., 2022) were subsequently linked to eligible institutions
according to the transformative agreement data. In the following, the
steps are described in detail.

@@

Comparing Crossref license coverage with OpenAlex, which re-uses open
access evidence from Unpaywall, a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), highlighted ongoing challenges in identifying
hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). For the purpose of this
study, 742,369 articles under CC license were retrieved using Crossref,
while 950,260 articles were tagged as ``hybrid'' according to the
OpenAlex November 2023 release, which was used throughout this study.
The largest differences concerned articles published between 2018 and
2020. In 2022, however, Crossref and OpenAlex open access numbers differ
only slightly (249,511 records using Crossref vs.~255,344 in OpenAlex).
Notable differences could be observed among some publishers that
presumably did not provide CC license information to Crossref, including
AIP Publishing, the American Physiological Society, Emerald, and the
Royal Society. Crossref license metadata was more complete with regard
to the articles published by Wiley and the American Chemical Society.
Finally, inconsistent open access status information in previous
OpenAlex versions was observed (Jahn et al., 2023). After reporting this
to OpenAlex, fixing this issue was still ongoing according to the
release notes, which might also explain part of the discrepancy.

@@

Throughout this mostly automated data gathering and analysis process,
Tidyverse tools (Wickham et al., 2019) for the R programming language (R
Core Team, 2020) were used. The resulting data are openly available
through an R data package, hoaddata (Jahn, 2023). Following Marwick et
al. (2018), hoaddata contains not only the datasets used in the data
analysis. It also includes code used to compile the data by connecting
it to the cloud-based Google Big Query data warehouse, where scholarly
big data from Crossref, OpenAlex and Unpaywall were imported, using
bigrquery (Wickham \& Bryan, 2020). To increase computational
reproducibility, data aggregation through hoaddata was automatically
carried out using GitHub Actions, a continuous integration service.

@@

Between 2018 and 2022, 11,189 out of 12,857 hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 of the 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles. Overall, 394 transformative
agreements recorded at least one open access article from an eligible
institution.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). Simultaneously,
the total article volume of the investigated journals rose from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable increase in open
access provision through other means, presumably publication fees being
not invoiced through transformative agreements, from 4.1\% (n = 62,625)
in 2018 to 6.3\% (n = 105,896).

@@

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, started offering open access
agreements for its hybrid journal portfolio to selected consortia such
as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish
FinELib consortium in 2015 under the name Springer Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
Prior to this, Springer had some pilot agreements with a small number of
institutions, including the University of Göttingen (Schmidt \& Shearer,
2012). However, the Springer Compact agreements were not included in the
data as they concluded before the start of the transformative agreement
data collection in June 2021. Nonetheless, the results demonstrate the
importance of agreements for Springer Nature's hybrid open access
business over the past five years (Figure 2B). In 2022, 66\% (n =
34,725) of open access in Springer Nature hybrid journals was enabled
through transformative agreements. In the same year, 70\% (n = 37,316)
of Wiley's open access articles could be linked to transformative
agreements in 2022. By contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

@@

\caption{\label{tab:subject_summary_table}Hybrid open access through transformative agreements by journal subject 2018-2022}
\centering
\begin{tabular}[t]{lrlrlrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Hybrid journals} & \multicolumn{2}{c}{Articles} & \multicolumn{2}{c}{OA articles} & \multicolumn{2}{c}{TA OA articles} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Journal subject & Total & \% & Total & \% & Total & \% & Total & \%\\
\midrule
Health Sciences & 2,342 & 22.5 & 1,998,045 & 28 & 199,265 & 27.7 & 81,913 & 25.7\\
Life Sciences & 1,399 & 13.4 & 1,080,346 & 15.1 & 133,526 & 18.6 & 48,570 & 15.2\\
Physical Sciences & 2,693 & 25.9 & 3,111,711 & 43.6 & 247,515 & 34.4 & 110,933 & 34.8\\
Social Sciences & 3,967 & 38.1 & 953,084 & 13.3 & 138,388 & 19.3 & 77,496 & 24.3\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018 and 2022. In particular, Social
Sciences including Arts and Humanties journals accounted for the
strongest growth in the five-years period from 6.4\% (n = 8,361) to 23\%
(n = 51,938), followed by the Life Sciences from 7.6\% (n = 15,003) to
18\% (n = 39,494) , Health Sciences from 5.3\% (n = 18,279) to 16\% (n =
63,089) and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n =
85,428). Growth in Social Sciences category can be largely attributed to
transformative agreements. In 2022, two-thirds of open access articles
(67\%, n = 34,759) were published by the first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B).

@@

Figure \ref{fig:subject_panel}C shows that this trend was consistent
across hybrid journals belonging to the ASJC Social Sciences category.
In 2022, 25\% of Social Sciences journals provided open access to at
least every fourth article exclusively through transformative
agreements. However, hybrid open access through transformative
agreements played a comparably lesser role in Life Sciences and Health
Sciences. In these two subject areas, only about half of the open access
articles can be linked to these agreements, both overall and on a median
average across journals. In contrast, the majority of Physical Science
journals show an increase of open access through transformative
agreements compared to other options to publish open access in hybrid
journals.

@@

Between 2018 and 2022, high-income countries almost exclusively
dominated hybrid open access publishing through transformative
agreements. During this period, first authors affiliated with
institutions from Organisation for Economic Co-operation and Development
(OECD) member countries published 602,050 open access articles in hybrid
journals, representing 81\% of the investigated open access articles.
This disparity between OECD nations and other countries becomes even
more evident when considering open access through transformative
agreements, as 310,712 of 328,957, or 94\% of open access articles were
associated with such agreements.

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by country, comparing the OECD area with the BRICS, an
intergovernmental organisation, which comprised Brazil, Russia, India,
China, and South Africa as of 2022. The residual category ``Other''
includes the remaining countries. Full counting was applied to account
for multiple country affiliations (Hottenrott et al., 2021). From 2018
to 2022, the proportion of open access in hybrid journals increased from
6.1\% in 2018 to 26\% in 2022. In contrast, BRICS recorded a low uptake,
moderately growing from 1.6\% in 2018 to 3.7\% in 2022.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares the total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (268,965 articles) and India (87,428
articles). In contrast, Western and Northern European countries
published a considerably high number of open access articles.
Particularly, Scandinavian countries, the Netherlands and Austria
recorded above-average open access shares, as indicated by the linear
trend line. As shown in Figure \ref{fig:country_patch}C, transformative
agreements contributed to these market positions. Interestingly, the
United States had a notable open access market share of 15\%, although
transformative agreements contributed to a lesser extent. Similarly,
China's open access market share of 7.2\% in 2022 was comparable to that
of the Netherlands, which was 7.1\%.

@@

Since 2021, there has been a general trend towards hybrid open access
among many high-income countries, driven primarily by transformative
agreements. However, the proliferation of transformative agreements
differed across these countries. \# For instance, Germany successfully
negotiated an agreement with Elsevier not until 2023. Additionally,
publication limits or eligibility criteria for institutions and article
types may explain why even countries with widespread agreement
implementation such as Sweden or the Netherlands did not achieve 100\%
hybrid open access. Interestingly, in Japan and the United States,
options other than transformative agreements were the main drivers for
the increase in hybrid open access. Once again, the graph highlights
countries with low hybrid open access, particularly non-OECD countries,
where only a few or no agreements were in place.

@@

The primary aim of this study was to assess the uptake of open access in
hybrid journals included in transformative agreements, which were
introduced as a temporal means to support the transition of
subscription-based academic publishing to full open access. This study
presents a novel approach based on open data, which leverages metadata
on over 700 agreements and nine million journal articles to estimate the
extent to which transformation agreements contribute to the transition
of this journal business model. The results highlight a strong growth in
open access between 2018 and 2022, driven by an increasing number of
institutions that had transformative agreements in place. However, the
majority of research literature published in hybrid journals in this
five-year period remained behind publisher paywalls. Growth in the
adoption of open access in hybrid journals, in particular through
transformative agreements, can be largely attributed to three large
commercial publishers -- Elsevier, Springer Nature, and Wiley -- but
varies substantially across journals, publishers, disciplines, and
country affiliations. Despite the limitations of the data, the findings
indicate that the current level of implementation of transformative
agreements is insufficient to bring about a large-scale transition to
fully open access.

@@

Moreover, this study presents varying levels of open access uptake
through transformative agreements across journals, which can be
attributed to the alignment of authors' affiliations and the
availability of such agreements at their institutions. In line with
previous research findings (Butler et al., 2023; Jahn et al., 2022;
Wenaas, 2022), high uptake rates were observed across hybrid journals in
Social Sciences and Humanities. However, it is important to emphasise
that these hybrid journals do not encompass the entire field. For
example, Khanna et al. (2022) found that approximately 60\% of journals
utilizing the open-source publishing platform Open Journal Systems (OJS)
fall within the Social Sciences and Humanities. In these fields, the
majority of full open access journals are so-called ``Diamond OA
journals'', which do not charge publication fees (Bosman, Frantsvåg, et
al., 2021). Rather, this result can be more accurately attributed to the
substantial proportion of authors from high-income countries who publish
in these hybrid journals included in transformative agreements,
particularly when journals are of local relevance, such as by belonging
to national societies or regional research scope.

@@

This large-scale study provides the first empirical evidence of the
influence of transformative agreements on the transition of hybrid
journals to full open access. However, several limitations needs to be
acknowledged. From a data perspective, estimations of open access
through transformative agreements were established by linking first
author affiliations with publicly available agreement data from
cOAlition S, and not through invoicing data, which is usually not
shared. Moreover, this study is unable to account for the various types
of transformative agreements, particularly with regard to article types
and caps that limit the number of open access articles covered.
Furthermore, assessing the quality of the OpenAlex and Crossref data
used, particularly in terms of affiliations and article types, by
comparing them with established bibliometric databases such as Scopus
and Web of Science was beyond the scope of this analysis. Overall, the
methodology is designed in such a way that the results underestimate
rather than overestimate the adoption of open access through
transformative agreements. In addition, the data used in this study are
openly available, along with the code used for this analysis.

Additionally, it must be noted that the study period was significantly
impacted by the COVID-19 pandemic, which led to an unprecedented number
of publications and a reduction in international collaboration
(Aviv-Reuven \& Rosenfeld, 2021), which could explain the observed
contrasting developments in OECD and BRICS countries. Even before the
pandemic, growth in publications in Europe was only due to
internationally co-authored journal articles (Kwiek, 2021). Likewise,
inflows from China to the US and European countries already declined by
2020 (Zhao et al., 2023). Furthermore, the study design did not consider
emerging publication practices such as preprints (Fraser et al., 2021)
and special issues (Hanson et al., 2023), which have grown rapidly since
2020. Lastly, it should be emphasised that the study did not address
financial shifts between subscriptions spending and open access payments
while analysing hybrid open access through transformative agreements due
to a lack of data on expenditures.

This study allows for multiple strands of further research. One is to
complement this large-scale study with more specific evidence from
individual countries or subjects, particularly those with low hybrid
open access rates. Incorporating full open access and subscription-based
journals, as well as considering global trends in scholarly migration
and collaboration could also be promising. Financial studies could build
on the study design and include subscription and open access expenditure
to assess the cost-effectiveness of transformative agreements, in
particular whether transformative agreements can integrate the
substantial amounts of individual payments for publication fees (Butler
et al., 2023; Wenaas, 2022), as well as potential changes in authors'
behaviour following the introduction of these agreements (Schmal, 2024).

@@

In summary, this study provides empirical insights into the development
of hybrid open access following the introduction of transformative
agreements. These results are important for both researchers and
stakeholders engaged in negotiating and evaluating these agreements. The
presented approach relies on open data, which enables follow-up studies
and open access monitoring activities to further explore the role of
transformative agreements in transitioning academic publishing to full
open access.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Brainard_2023}{}}%
Brainard, J. (2023). {``Transformative''} journals get booted for
switching to open access too slowly. \emph{Science}.
\url{https://doi.org/10.1126/science.adj3282}

@@

\leavevmode\vadjust pre{\hypertarget{ref-ifla}{}}%
Campbell, C., Dér, Á., Geschuhn, K., \& Valente, A. (2022). How are
transformative agreements transforming libraries? \emph{87th IFLA World
Library and Information Congress (WLIC) / 2022 in Dublin, Ireland}.
\url{https://repository.ifla.org/handle/123456789/1973}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Geschuhn_2017}{}}%
Geschuhn, K., \& Stone, G. (2017). It's the workflows, stupid! What is
required to make {``offsetting''} work for the open access transition.
\emph{Insights the {UKSG} Journal}, \emph{30}(3), 103--114.
\url{https://doi.org/10.1629/uksg.391}
